On the effects of partying on next-day sports performance

### 2017-06-03 writing/draft.tex
Read-through with Eric in Kennedy.

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
Anecdotally, it has been reported that professional sports teams party on 
the road, sometimes to the detriment of the game.
Within the NBA, the player J.R. Smith underwent a substantial improvement
in performance metrics after being traded from New York to Cleveland; when an interviewer
asked what contributed to this bump in statistics, he infamously replied
that in New York he would have been out partying, but because there's no nightlife
in Cleveland, he ends up hitting the gym.\cite{price,ley}
Matt Barnes recently complained that ``there's no nightlife in Utah'' when 
asked about playing Jazz in the second round of NBA playoffs in 2017.\cite{cestone}
A few interviews with former professional NBA athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

@@

\section{Previous Research}
There has been an attempt to answer this question before.\cite{ezekowitz} 
However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, their sample size was much smaller. 
Finally, there was no methodology used behind the selection of party cities. The author does conclude
that teams systematically play worse on Sunday afternoons when on the road,
regardless of cities.

Studies have also been carried out on statistical ranking of the players
most notorious for pursuing active nightlife in the NBA.\cite{chase} The study
does conclude that players perform worse on Sundays when compared with other days
as measured by the ratio of shots scored to shots taken.

@@

\paragraph{Two approaches: a discrete indicator and a continuous measure of nightlife}
Because famous athletes often enjoy a celebrity lifestyle, using conventional statistics
such as number of drinking establishments in each city does not work; athletes instead
tend to visit high end nightclubs. We again turned to player interviews for ideas,
and found two major insights. From our conversations, 
we obtained a listing of cities which were anecdotally
mentioned as having active night-life. Between players, the rank ordering of these
was consistent to the extent that Los Angeles and New York city were top
offenders. This motivates using a binary indicator for party in these two cities. 
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a
simple count of the number of musicians in the county in which the stadium resides. 
We pursue both approaches.

@@

\paragraph{Fatigue from jetlag}
What about jetlag? Prolonged air travel has in general been shown
to have a negative effect on sports performance.\cite{leeandgalvez}
Previous studies have shown that jetlag is a strong enough effect
to erase home field advantage in major league baseball, for example.\cite{songetal}

\paragraph{The problem of an unknown travel schedule} In NBA, we have the difficulty of not observing when teams travel. E.g. if a particular team  plays
in Los Angeles on Friday and another game in New York the following day, we don't
know whether the team traveled on Friday evening or Saturday morning and hence don't know
which city they ``visited'' on Friday night. Heuristics for
travel schedules can be made, but it is hard to obtain reliable team-travel data.
In MLB, on the other hand, each match-up is actually a series of games. E.g.
if the Dodgers play the Giants in San Francisco, they will play several games spread
across several days, each in the same city; in these cases, we can infer that the team
is in fact spending the night in said city.

@@

We then merge this in with our lines data, taking care to do so according to each season-team's last game location such that the feature captures the proper
party effect. In addition, we lag our BLS data by one year for two reasons:
realistically, player expectations of which cities have the most active night-life may not
update in real-time and also doing so allows us to create a betting model, i.e.
we don't rely on any features we can't obtain before making a game-time prediction. 
To this data, we've added features such as 
the number of days since last game, the last game location, and the travel distance.

\paragraph{Minutes data}
We also gathered minutes data from \href{http://www.espn.com/nba/scoreboard/_/date/}{ESPN}.
The minutes data also contain information on the number of free throws, field goals, and three pointers made and attempted by each player in each game.

@@

\subsection{Method: Defining what it means for a city to have nightlife}
We have a free parameter: the time-window which we believe the hangover effect
to last. We consider games which have been played at most 24 hours ago, and take care to ignore
teams playing in their own city, i.e. if a team's last game was played at home  we say that they were not exposed to an exogenous shock in nightlife.

\paragraph{Discrete indicator}
Our first sanity check was to define an indicator variable for when a team plays in 
Los Angeles or New York within 24 hours of their current game, we indicate one, else zero. Anecdotally, these are mentioned as hot-spots for
nightlife.

\paragraph{Continuous measure}
We also defined a continuous measure of nightlife:
if the last game was played at most 24 hours ago, use the log of the average number
of musicians in last game location, else emit zero. Intuitively, this captures
how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.

@@

We mention that it's not surprising that features describing jet-lag and recovery time
are not significant: such salient information has already been taken into account in
setting the spread.
In the next section, we explore which specific areas of gameplay are affected.

@@

\subsection{Robustness check}
As a robustness check, we verify that this phenomena persists in an entirely separate
sport with entirely different players. To that end, we look toward
Major League Baseball, which is also well set up for our quasi-experiment since
there are many games played back-to-back in the season (more in fact, than NBA).
There are also anecdotes of players celebrating with alcohol in public.\cite{miller,campbell}

@@

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, The 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting for the bookmaker's cut. E.g. in the     early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k.}
  \label{bettingperf}
\end{figure}

>>>>

\section{Introduction}
It's not unreasonable to think that game performance can be affected 
(in part) by what takes place off the court. 
There are numerous reports of professional sports teams partying on 
the road, sometimes to the detriment of the game.
Within the NBA, J.R. Smith's performance metrics substantially improved 
 after being traded from New York to Cleveland; when an interviewer
asked what contributed to this bump in statistics, he infamously replied
that in New York he would have been out partying, but because there's no nightlife
in Cleveland, he ends up hitting the gym.\cite{price,ley}
Matt Barnes recently complained that ``there's no nightlife in Utah'' when 
asked about playing Jazz in the second round of NBA playoffs in 2017.\cite{cestone}
A few interviews with former professional NBA athletes revealed that location
matters more than day of the week, i.e. parties can happen any day of the week,
but they are more likely to occur in select cities.

@@

\section{Previous Research}
There has been an attempt to answer this question before.\cite{ezekowitz} 
However, the study only
looks at performance in party cities, as opposed to subsequent day performance (i.e. the study does not consider a delayed hangover effect). Moreover, their sample size was much smaller. 
Finally, there was no methodology used behind the selection of party cities. The author does conclude
that teams systematically play worse on Sunday afternoons when on the road,
regardless of the city.

Studies have also been carried out on statistical ranking of the players
most notorious for pursuing active nightlife in the NBA.\cite{chase} The study
 concludes that players perform worse on Sundays when compared with other days, as measured by the ratio of shots scored to shots taken.

@@

\paragraph{Two approaches: a discrete indicator and a continuous measure of nightlife}
Because famous athletes often enjoy a celebrity lifestyle, using conventional statistics
such as number of drinking establishments in each city does not work; athletes instead
tend to visit high end nightclubs. We again turned to player interviews for ideas,
and found two major insights. From our conversations, 
we obtained a listing of cities which were anecdotally
mentioned as having active night-life. Between players, the rank ordering of these
was consistent to the extent that Los Angeles and New York city were top
offenders. This motivates using a binary indicator for party in these two cities. 
We also found that athletes tend to party with other A-list celebrities,
particularly musicians. This motivated a continuous measure of nightlife, based on a
simple count of the number of musicians in the county in which the stadium resides. 
This captures
how much nightlife there is in a city since musical performances are correlated
with parties, if not an essential feature.
We pursue both approaches.

@@

\paragraph{Fatigue from jetlag}
What about jetlag? Prolonged air travel has in general been shown
to have a negative effect on sports performance.\cite{leeandgalvez}
Previous studies have shown that jetlag is a strong enough effect
to erase home field advantage in major league baseball.\cite{songetal}

\paragraph{The problem of an unknown travel schedule} In NBA, we have the difficulty of not observing when teams travel. E.g. if a particular team  plays
in Los Angeles on Friday and another game in Boston the following day, we don't
know whether the team traveled on Friday evening or Saturday morning and hence don't know
which city they ``visited'' on Friday night. Heuristics for
travel schedules can be made, but it is hard to obtain reliable team-travel data.
In MLB, on the other hand, each match-up is actually a series of games. E.g.
if the Dodgers play the Giants in San Francisco, they will play several games spread
across several days, each in the same city; in these cases, we can infer that the team
is in fact spending the night in said city.

@@

We then merge this in with our lines data, taking care to do so according to each team's last game location such that the feature captures the proper
party effect. In addition, we lag our BLS data by one year for two reasons:
realistically, player expectations of which cities have the most active night-life may not
update in real-time and also doing so allows us to create a betting model, i.e.
we don't rely on any features we can't obtain before making a game-time prediction. 
To this data, we've added features such as 
the number of days since last game, the last game location, and the travel distance.

\paragraph{Minutes data}
We also gathered minutes data from \href{http://www.espn.com/nba/scoreboard/_/date/}{ESPN}.
The minutes data contains information on the number of free throws, field goals, and three pointers made and attempted by each player in each game.

@@

\subsection{Method: Defining what it means for a city to have nightlife}
We consider games which have been played at most 24 hours ago, and take care to ignore
teams playing in their own city, i.e. if a team's last game was played at home  we say that they were not exposed to an exogenous shock in nightlife.

\paragraph{Discrete indicator}
Our first sanity check was to define an indicator variable for when a team plays in 
Los Angeles or New York within 24 hours of their current game, we indicate one, else zero.

\paragraph{Continuous measure}
We also defined a continuous measure of nightlife:
if the last game was played at most 24 hours ago, use the log of the average number
of musicians in last game location, else emit zero.

@@

The features describing jet-lag and recovery time are not significant, but this is to be
expected: such salient information has already been taken into account in setting the spread.
In the next section, we explore which specific areas of gameplay are affected.

@@

\subsection{Robustness check}
As a robustness check, we verify that this phenomena persists in an entirely separate
sport. To that end, we look toward
Major League Baseball, which is also well set up for our quasi-experiment since
there are many games played back-to-back in the season (more in fact, than NBA).
There are also anecdotes of players celebrating with alcohol in public.\cite{miller,campbell}

@@

\subsubsection{Betting performance}
We also validate our MLB model by seeing how well we can perform against the
market.
\begin{figure}[!h]
  \centering
  \includegraphics{mlb_bets_by_season}
  \caption{Betting performance by season. In each season, the 
    circled point represents the worst out of pocket expense incurred.
    We first train a model on historical data then use said model to predict on ``current year''
    data, plotting profits as a function of time over the ``current'' season. For each relevant
    game, we place a bet of \$100 if the expected value of the bet is positive after accounting for the bookmaker's cut. E.g. in the     early 2016 season,
    at its worst we are out of pocket \$750 dollars. By season end, we have     recouped all of
    our initial investment plus an additional \$11.5k. We are profitable in each season, and mention that 2015 is the ``worst'', but we still come out ahead.}
  \label{bettingperf}
\end{figure}
